---
title: 'Weel 2 - 2022-09-19: Economic Modeling'
author: Joshua Doolan
date: '2022-09-19'
slug: []
categories: []
tags: []
---

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(tidyverse)

# Import Data
natl_unemp = read_csv("Section Data/unemployment_national_quarterly_final.csv")
CPI_df = read_csv("Section Data/CPI_monthly.csv")
GDP_df = read_csv("Section Data/GDP_quarterly.csv")
RDI_df = read_csv("Section Data/RDI_quarterly.csv")
house_pop = read_csv("Section Data/house_popvote_seats.csv")
```

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
# Import Data
midterm_year = (natl_unemp$year %% 2 == 0) & (natl_unemp$year %% 4 != 0)
natl_unemp$midterm_year = midterm_year
unemp_diff = natl_unemp %>% 
  filter((midterm_year == TRUE & (quarter_yr == 3 | quarter_yr == 4))) %>% group_by(year) %>% mutate(unemp_diff = UNRATE - lag(UNRATE)) %>% filter(quarter_yr == 4) %>% select(year, UNRATE, unemp_diff)

natl_df = left_join(unemp_diff, house_pop %>% select(year, H_incumbent_party_majorvote_pct), by = "year")
```

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
CPI_df$year = as.numeric(format(CPI_df$DATE, "%Y"))
CPI_df$month = as.numeric(format(CPI_df$DATE, "%m"))
CPI_data = CPI_df %>% filter(year %% 2 == 0 & year %% 4 != 0 & month %in% c(7, 10)) %>% group_by(year) %>% mutate(CPI_diff = CPIAUCSL - lag(CPIAUCSL)) %>% filter(month == 10) %>% select(year, CPI_diff)
natl_df = left_join(natl_df, CPI_data, by = "year")
```

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
GDP_data = GDP_df %>% 
  filter(year %% 2 == 0 & year %% 4 != 0 & quarter_yr == 4) %>% select(year, GDP_growth_pct)
natl_df = left_join(natl_df, GDP_data, by = "year")
RDI_data = RDI_df %>% 
  filter(year %% 2 == 0 & year %% 4 != 0 & quarter_yr == 4) %>% select(year, DSPIC_change_pct)
natl_df = left_join(natl_df, RDI_data, by = "year")
```

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
natl_lm_unemp = lm(H_incumbent_party_majorvote_pct ~ UNRATE, data = natl_df)
natl_lm_undiff = lm(H_incumbent_party_majorvote_pct ~ unemp_diff, data = natl_df)
natl_lm_CPI = lm(H_incumbent_party_majorvote_pct ~ CPI_diff, data = natl_df)
natl_lm_GDP = lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct, data = natl_df)
natl_lm_RDI = lm(H_incumbent_party_majorvote_pct ~ DSPIC_change_pct, data = natl_df)
```

```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
predict(natl_lm_unemp, data.frame(UNRATE = 3.6))
predict(natl_lm_undiff, data.frame(unemp_diff = -0.4))
predict(natl_lm_CPI, data.frame(CPI_diff = 8.6))
predict(natl_lm_GDP, data.frame(GDP_growth_pct = 2.03))
predict(natl_lm_RDI, data.frame(DSPIC_change_pct = -33.3))
```

Hello everyone! For this week's blog, we're going to be taking a look at how various different economic indicators (specifically GDP, RDI, CPI, and Unemployment) produce different outcomes for the 2022 midterm elections. 

Before diving into the details of these different models, I wanted to mention, if you're interested in seeing some of the "scratch work" for how I come to these analyses, please feel free to view my weekly notebooks on my Github (this week's can be found here [here](https://github.com/joshua-doolan/GOV1347-Blog-Posts-F22/blob/main/content/post/2022-09-19-weel-2-2022-09-19-economic-modeling/week2_R.ipynb)). These contain
