---
title: Week 4 - 10/3/22 - Incumbency & Experts
author: Joshua Doolan
date: '2022-10-03'
slug: []
categories: []
tags: []
---

# Model Update
This week, I will be taking a different approach to vote-share prediction than before. This will be my first week incorporating district-level predictions. District-level predictions all for more nuance, but suffer from a lack of commonly accessible data (a large issue for linear regression models). To begin this district analysis, I will focus on incumbency and expert predictions.

As always, scratch work can be found [here](https://github.com/joshua-doolan/GOV1347-Blog-Posts-F22/blob/main/content/post/2022-10-03-week-4-10-3-22-incumbency-experts/week4_py.ipynb).

## Incumbency
Incumbency has been widely shown to be a large predictor of electoral outcomes. As such, predicting all race outcomes based on incumbency alone would lead to a boring task, as by predicting 100\% incumbent wins re-election will result in a correct prediction 90\% of the time. Therefore, I made two modifications to the set up. First, I will predict vote-share as opposed to probability of winning (as vote share can have much larger fluctuations) and second I will only predict competitive races, by definition lowering the incumbent's chance of winning (by definition of a competitive race). Importantly for incumbency, I will define it as two separate indicator variables, one if the democrat is an incumbent, and the second variable if the republican is the incumbent. This is a useful definition (as opposed to one variable, which is 1 for Dem and -1 for Rep, for example) for linear regression for multiple reasons.
- Asymetric advantage: this allows Republican and Democratic incumbents to be treated different.
- Multiple incumbents: when redistricting, if the district merges between opposite parties, we can handle the advantages.
- No incumbents: In the case of no incumbents, the weight goes to 0.
However, by considering only candidate incumbency, I am losing valuable party information (e.g. if a longtime representative retires, my model gives no advantage to their party). However, I hope to capture this information via the expert predictions. 

## Expert Predictions
To gather underlying information on district partisan preferences (which we are predicting), I decided to use the expert ratings as well. This is particularly important in this redistricting cycle, as using prior performance of a district may be inaccurate given the changing of district boundaries. Accordingly, a solution for this may be to use district similarity as discussed in section (similarity to historical districts) but for the time being, I will stick with the base linear model. The second challenge for this is the inconsistent expert information. One expert may make a prediction for a district that another expert does not. Since linear regression has issues with missing data, I instead used the average rating of all the experts who made prediction for a given race.

## District Data / Modeling Choices

Because of the redistricting changes mentioned above, instead of creating a linear model for each district, and perhaps incorporating some outside information via weighting, I instead made one competitive race model, solely using the factors above. This resulted in a R-squared of .72, which performed better than national level predictions.

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library("tidyverse")
competitive_races = read_csv("competitive_races.csv")
rating_lm = lm(RepVotesMajorPercent ~ avg_rating + `Rep Inc` + `Dem Inc`,data=competitive_races)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(rating_lm)
```

Notably, the expert rating factor outweighs incumbency, likely because the experts already include incumbency in their ratings. However, in a nice outcome, we see a nearly symmetric weight given to Republican and Democratic incumbency, suggesting that the power of incumbency is symmetric.

# Blog Extension -- Expert Performance in 2018

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
expert_means = read_csv("expert_means_2018.csv")
```

Moving on to analyzing the results of the experts in the 2018 midterms, I looked at Daily Kos, 538, and the Cook predictions. Notably, all of these are on a 7 point scale, where 1 means the expert believes the democrat is favored, 4 means it is a pure toss up, and 7 means that the race favors the Republican.

What we find is that the Democrats performed better than the (non-538) experts predicted. This is notable in a couple areas. First, when rating a race as a toss-up, for both Cook and Daily Kos, Democrats actally won an increased vote share. 538 however, was more correct on their tossup, as Democrats and Rebulicans tied on average. Second, we can see that the chart is not symmetric: The bars on the left side are longer than the bars on the right side, showing that overall democrats had a relative outperformance is estimation. Finally, we see that even 538 is not asymmetric, as when 538 favored a democrat (1), the Democrat won on average by 15 points, but when they favored the Republican (7), the Republican won by only 12 points.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=expert_means, aes(fill=expert, x=rating, y=avg)) +  geom_bar(position="dodge", stat="identity")
```

Overall, this analysis shows that while we may inevitably expect experts to be occasionally wrong on the individual race, we find that experts can also be quite wrong in the aggregate as well, showing that even the most in-depth analysis is subject to unforeseen swings.