---
title: Week 5 (10/11/22) - Air War
author: Joshua Doolan
date: '2022-10-11'
slug: []
categories: []
tags: []
---

# Model Update
This week, I will be predicting district-level Republican Two Party Vote Share from advertising data.

As always, scratch work can be found [here]https://github.com/joshua-doolan/GOV1347-Blog-Posts-F22/blob/main/content/post/2022-10-11-week-5-10-11-22-air-war/week5_py.ipynb).

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library("tidyverse")
ads_races = read_csv("Section data/train_ads.csv")
```

# Advertising Data

This week, I take a look at two possible predictors for advertising on the district level: absolute advertising spend advantage. In theory, these are both measures of the same underlying data (both rely on $ spent), but have different potential modeling effects:

- Absolute Advertising Spend Advantage: this is the \$ more that Republicans spent on ads than Democrats. In theory, this provides more information than ad spend proportion, as we can imagine in low-ad spend races that the proportion could be very high, but the absolute advantage low, and the average voter may have seen no ads. However, this is potentially difficult to use in a linear model, as the large outliers will have outsize effects on the parameters.
- Scaled Advertising Spend Proportion (1 if only Republicans spend, -1 if only Democrats spend, 0 if Dem Spend = GOP Spend): In theory, the vast majority of races will have smaller spending than the extremes, so this should capture that information. These are scaled to avoid the trap of having full democrat spend at 0 causing the linear model from being unable to capture this data, and to differentiate from races with no spending.

# Models

## Model 1: Baseline (Expert Ratings + Incumbency)

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
rating_lm = lm(RepVotesMajorPercent ~ avg_rating + `Rep Inc` + `Dem Inc`,data=ads_races)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(rating_lm)
```

Included as reference, we see a $R^2$ of 0.71, with each variable having very low p-values, showing that this is a dense baseline (no unnecessary variables).

## Model 2: Absolute Advertising Advantage

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
ads_lm = lm(RepVotesMajorPercent ~ avg_rating + `Rep Inc` + `Dem Inc` + `Rep Spend Advantage`,data=ads_races)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(ads_lm)
```

We see here that the spending advantage has led to a very low ($10^{-8}$) coefficient for advertising spend advantage. In reality, this means that in order to see a 1\% increase in vote share, a spending advatnage of $100M is needed (for an individual race). This would seem to indicate that advertising advantage is not very predictive. While our R^2 barely improves to 0.72, this indicates that for most every congressional race, the most amounts of spending only swing single digit votes.  

## Model 3: Advertising Proportion

```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
ads_prop = lm(RepVotesMajorPercent ~ avg_rating + `Rep Inc` + `Dem Inc` + `Rep Spend Proportion`,data=ads_races)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(ads_prop)
```

We see here the same effect on R^2, but note that we see a negative coefficient on advertising spending. This seems to indicate that there is so much noise (also indicated by the p value of 0.89), that this is likely a spurious correlation, and provides less information on the effects on ads, even though the coefficient is caled to a more reasonable standard.

Accordingly, I would exclude advertising spending from future models, but would be interested in studying weighting by recency. However, since most spending occurs at the end of the race, I would imagine a recency weighting to have little effect, as the data is already essentially recently biased.

# Blog Extension 3: GPR

Gerber et al. estimate GPR by conducting telephone surveys asking if respondents had seen an ad (although whether they asked this question was randomized, so they only asked it for 1/3 of questions). Using this, they estimate the reach. In theory, to measure advertising effects in the digital age, you could estimate who was reached by a digital ad (by targeting them to specific people), and then measuring turnout post-facto using voting data. You could also run a survey where you show people an ad, and then ask aboutv= voting.